Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 BMC-Response-Round1.tex
Updates from ShareLaTeX

\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint is not intended to perform negative design. It is a generation algorithm, not only addressing a large array of design use-cases, but also able to generate good seed sequences, amenable to further optimization using other tools. For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform). We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.}

@@

\AnswerTODO[Yann]{Again, we do not feel the need to validate, or further analyze, the MultiDefect metrics, which has been a standard for multiple structure design. The time taken by the computational design is generally negligible compared to the time taken to experimentally validate the ultimate functionality of a design. What we believe to matter more to the community is the expressivity of the tool which, in this case, allows to address various use-cases, including some that could not be addressed before.
}
\Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%
\Answer{In the current implementation, it is only possible to constrain the global GC\% of the sequence. It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (eg  structures are already associated with individual target energies). Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version, should the concrete need arise.}
\Comment{
The section "problem statement" is not helpful enough for readers, both of algorithmic and biological people.
In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)
\\[1em]}%
\AnswerTODO[Yann]{We attempted to clarify it.
}
\Comment{
Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1. The description of the brief aim and summary of each step would be a good guide for readers.\\[1em]}%
\AnswerTODO[Yan]{
Add method overview
}

>>>>

\Answer{As we tried to make clear in our general comments of this response letter (as well as by appropriate adaptions in the manuscript), RNAredprint does not by itself perform \emph{negative} design. It is a generation algorithm for (positively) designing sequences with complex properties. In this way, it addresses a large range of design use-cases, but is also able to generate good seed sequences, amenable to further optimization using other tools. For this reason, we focused our validation on this potential for better initialization, and compared the quality of sequences generated by RNARedPrint (Boltzmann weighted) to RNABlueprint (uniform). We noticed that, while RNAredprint is not explicitly tailored to optimize MultiDefect, it nevertheless produces sequences that: 1) achieve better values; and 2) are amenable to better further optimizations than the uniform version implemented by RNABluePrint.}

@@

\Answer{For reasons already outlined above, we think that a sophisticated analyses of the MultiDefect metrics, which has been a standard for multiple structure design, should not be part of this work. Nevertheless, we report and discuss a breakdown of the measure into its two components.
The final suggestion seems more interesting for negative design, since in positive design, we can directly prescribe sequence characteristics.}

\Comment{(Is RNARedPrint able to produce, for example, the combination of two long stem loops which are high and low GC-content, respectively?)\\[1em]}%

\Answer{In the current implementation, it is only possible to constrain the global GC\% of the sequence. It is however possible to support this feature within our general framework, since one can associated individual objectives to structures (e.g.\ structures are already associated with individual target energies). Different stems can be expressed as different structures by "peeling the structure", so this feature could be added to a future version.}

\Comment{
The section "problem statement" is not helpful enough for readers, both of algorithmic and biological people.
In particular, it is difficult to comprehend the relationships between the variables due to insufficient descriptions and lack of definitions.
For example, they do not specify any relationship between R and F in this section while they should not be independent. (e.g., m should be equal or less than k?)}%

\Answer{We understand that the text was still hard to read and thus rewrote large parts of the first subsections in Methods to make it more accessible.}

\Comment{
Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1. The description of the brief aim and summary of each step would be a good guide for readers.}%
\Answer{We included additional explanations and took care to define these variables. We considered the incl further advantages.}

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

\revised{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function. Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules. One can thus sharply reduce the complexity of Algorithm~\ref{alg:pf} by restricting the precomputations to compatible assignments.}

@@

However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As can be shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp. Mat.~\ref{appsec:modelparameters} can be used to approximate the Turner energy model (correlation coefficient $R=0.99$), while greatly reducing the tree width, and thus the computational requirements of the whole method even for complex instances.

}

\subsection*{Effectively targeting Turner energies using multidimensionnal sampling}
\revised{

We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp.\ Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

We considered a benchmark of six sets of target structures described in \citep{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design.
Using \ourprog{}, for
each instance of the benchmark, we generated at least 1\,000 seed sequences with similar
energies for all target structures. For this purpose, we determine good common target energies that
can be succesfully targeted for all single target structures simultaneously.
Generally, we targeted 60\% \GCb\%.  We compared the
\Obj{} value of these sequences against that of seed sequences,
uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
\Obj{} function. At each step, the local search resamples (uniformly
at random) the positions of a randomly selected component in the base
pair dependency graph, accepting the modification only if it results
in a gain. We performed 500 greedy descent steps in the case of
pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
  LE80}.

>>>>

\revised{Proposition~\ref{prop:bijection} also strongly impacts the complexity of computing the partition function. Indeed it implies that, among the $4^k$ possible assignments of nucleotides to $k$ connected positions (in the base-pair dependency graph), at most $2^k$ are compatible with base pairing rules. One can thus sharply reduce the complexity of Algorithm~\ref{alg:pf} by restricting the precomputations to compatible assignments.

For a discussion on the implications of our hardness results beyond exact counting see Supp.\ Mat.~\ref{appsec:approximate-counting}.
}

@@

However, while in principle feasible, sampling in a Boltzmann distribution directly using the Turner energy model may induce extreme computational demands, with tree-widths scaling at least as large as the number of nucleotides in the largest loop. Fortunately, we found that intricacy of the Turner energy model can be circumvented with minimal loss of precision by using a simpler stacking energy model. As shown in Figure~\ref{fig:correlation}, a simple stacking energy model, further described in Supp. Mat.~\ref{appsec:modelparameters}, can be used to well approximate the Turner energy model (correlation coefficient $R=0.99$) for our purpose. This greatly reduces the treewidth, and thus the computational requirements of the whole method even for complex instances.
}

\subsection*{Effectively targeting Turner energies using multidimensionnal sampling}
\revised{
We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp.\ Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

We considered a benchmark of six sets of target structures described in \citep{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design.
Using \revised{multi-dimensional Boltzmann sampling based on} \ourprog{}, for
each instance of the benchmark, we generated at least 1\,000 seed sequences with similar energies for all target structures. 
For this purpose, we determine good common target energies that
can be successfully targeted for all single target structures simultaneously.
Generally, we targeted 60\% \GCb\%.
\revised{As described in Section~\ref{sec:multiBoltzmann}, to find sequences fulfilling our energy and \GCb\% criteria, we iteratively sampled 10\,000 sequences and kept admissible ones while optimizing the weights. In practice, this sampling operation performs quite fast (see \ref{appsec:run-times}) and only little iterations were needed for most of the benchmark inputs as reported in  \ref{appsec:effectivity-seed-generation}.}
We compared the
\Obj{} value of \revised{the derived} sequences against that of seed sequences,
uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
\Obj{} function. At each step, the local search resamples (uniformly
at random) the positions of a randomly selected component in the base
pair dependency graph, accepting the modification only if it results
in a gain. We performed 500 greedy descent steps in the case of
pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
  LE80}.
